---
title: "project 2"
author: "Michael"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1: Synopsis
The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1. Health (injuries and fatalities) 
2. Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
## import relevant R library
the following R library shall be imported for processing and analysis of data and setting working directory
```{r}
setwd("./")
path<-"./"
if(!require("tidyverse")){
  install.packages("tidyverse")
  library(tidyverse)
}
if(!require("tidyr")){
  install.packages("tidyr")
  library(tidyr)
}

if(!require("dplyr")){
  install.packages("dplyr")
  library(dplyr)
}

if(!require("ggplot2")){
  install.packages("ggplot2")
  library(ggplot2)
}

if(!require("chron")){
  install.packages("chron")
  library(chron)
}

if(!require("MASS")){
  install.packages("MASS")
  library(MASS)
}

if(!require("reshape2")){
  install.packages("reshape2")
  library(reshape2)
}
if(!require("reshape")){
  install.packages("reshape")
  library(reshape)
}

```
## Data processing
### Importing data set

the dataset was downloaded from the url [download dataset][1],and imported using relevant R library

```{r,results='hide'}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!(file.exists("./Storm data.csv")))
{ download.file(url = url,
              destfile = paste(path, "Storm Data.csv", sep = "/"))}
if (!file.exists("./Storm Data.csv"))
{unzip(zipfile  = "Storm Data.zip")}
              
data<-read.csv("./Storm Data.csv")
head(data,n=6)
```

### selecting relevant data variable
```{r}
req_data <-data%>%select(c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP"))
#only use data where where fatalities and injury occurred
req_data<-req_data[(req_data$EVTYPE != "?" & 
             (req_data$INJURIES > 0 | req_data$FATALITIES > 0 | req_data$PROPDMG > 0 | req_data$CROPDMG > 0)), ]
```

### Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop damage cost.
```{r CorrectingExponents, results="hide"}
# Change all damage exponents to uppercase.
cols <- c("PROPDMGEXP", "CROPDMGEXP")

# Map property damage alphanumeric exponents to numeric values.
propdmgexp <- as.character(req_data[,"PROPDMGEXP"])
unique_propexp <- unique(propdmgexp)
cropdmgexp <- as.character(req_data[,"CROPDMGEXP"])
unique_Cropexp <- unique(cropdmgexp)
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)
# Map crop damage alphanumeric exponents to numeric values
cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)

propDmgKey_transformed <- propDmgKey[propdmgexp]
class(propDmgKey_transformed)
class(propDmgKey)

cropDmgKey_transformed <- cropDmgKey[cropdmgexp]
req_data <- mutate(req_data,PROPDMGEXP = propDmgKey_transformed)
req_data$PROPDMGEXP[is.na(req_data$PROPDMGEXP)] = 10^0 #replacing all na values in PROPDMGEXP column with 10^0
req_data[is.na(req_data$PROPDMGEXP),"PROPDMGEXP"]       # verify that all na values in the PROPDMGEXP column have been replaced

req_data <- mutate(req_data,CROPDMGEXP = cropDmgKey_transformed)
req_data$CROPDMGEXP[is.na(req_data$CROPDMGEXP)] = 10^0
req_data[is.na(req_data$CROPDMGEXP),"CROPDMGEXP"]

```

### 2.5: Making Economic Cost Columns

```{r }
req_data <- mutate(req_data,propCost = PROPDMG * PROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP,Total_Cost = (propCost + cropCost))
```


## grouping data by event types and summarising the data
```{r }
table_1<- req_data %>%
  group_by(EVTYPE) %>%
  summarise(FATALITIES = sum(FATALITIES),INJURIES = sum(INJURIES), PROPERTY_DAMAGES = sum(propCost), CROP_DAMAGES = sum(cropCost),Total_Cost = sum(Total_Cost), .groups = 'drop') %>%
  arrange(EVTYPE)%>%
  print(n=30)
# creating different table ordered by different session and selecting the top 20

table_fat<- table_1[order(-table_1$FATALITIES), ]
table_fat<-table_fat[1:20,]
table_inj<- table_1[order(-table_1$INJURIES), ]
table_inj<-table_inj[1:20,]
table_propdmg <- table_1[order(-table_1$PROPERTY_DAMAGES), ]
table_propdmg<-table_propdmg[1:20,]
table_cropdmg <- table_1[order(-table_1$CROP_DAMAGES), ]
table_cropdmg<-table_cropdmg[1:20,]
table_total <- table_1[order(-table_1$Total_Cost), ]
table_total<-table_total[1:20,]
```

### 3.1: Events that are Most Harmful to Population Health


## Including Plots
### chart for top 20 events causing the most fatalities
You can also embed plots, for example:
```{r fatalityChart}
png("fatality.png")
# Create chart
fatalityChart <- ggplot(table_fat, aes(x=EVTYPE, y=PROPERTY_DAMAGES))
# Plot data as bar chart
fatalityChart = fatalityChart + geom_bar(stat="identity", aes(fill=EVTYPE), position="dodge")
# Format y-axis scale and set y-axis label
fatalityChart = fatalityChart + ylab("Frequency Count") 
# Set x-axis label
fatalityChart = fatalityChart + xlab("Event Type") 
# Rotate x-axis tick labels 
fatalityChart = fatalityChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
fatalityChart = fatalityChart + ggtitle("Top 20 US Killers") + theme(plot.title = element_text(hjust = 0.5))
fatalityChart
dev.off()
```
### chart for top 20 events causing the most injuries
You can also embed plots, for example:
```{r injuryChart}
# Create chart 
png("injury.png")
injuryChart <- ggplot(table_inj, aes(x=EVTYPE, y=PROPERTY_DAMAGES))
# Plot data as bar chart
injuryChart = injuryChart + geom_bar(stat="identity", aes(fill=EVTYPE), position="dodge")
# Format y-axis scale and set y-axis label
injuryChart = injuryChart + ylab("Frequency Count") 
# Set x-axis label
injuryChart = injuryChart + xlab("Event Type") 
# Rotate x-axis tick labels 
injuryChart = injuryChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
injuryChart = injuryChart + ggtitle("Top 20 US Injuries") + theme(plot.title = element_text(hjust = 0.5))
injuryChart
dev.off()
```
### chart for top 20 events causing the most property damages
You can also embed plots, for example:
```{r propDmgChart}
# Create chart
png("propdmg.png")
propDmgChart <- ggplot(table_propdmg, aes(x=EVTYPE, y=PROPERTY_DAMAGES))
# Plot data as bar chart
propDmgChart = propDmgChart + geom_bar(stat="identity", aes(fill=EVTYPE), position="dodge")
# Format y-axis scale and set y-axis label
propDmgChart = propDmgChart + ylab("Frequency Count") 
# Set x-axis label
propDmgChart = propDmgChart + xlab("Event Type") 
# Rotate x-axis tick labels 
propDmgChart = propDmgChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
propDmgChart = propDmgChart + ggtitle("Top 20 US events causing property damages") + theme(plot.title = element_text(hjust = 0.5))
propDmgChart
dev.off()
```
### chart for top 20 events causing the most crop damages
You can also embed plots, for example:
```{r cropDmgChart}
# Create chart
png("cropdmg.png")
cropDmgChart <- ggplot(table_cropdmg, aes(x=EVTYPE, y=PROPERTY_DAMAGES))
# Plot data as bar chart
cropDmgChart = cropDmgChart + geom_bar(stat="identity", aes(fill=EVTYPE), position="dodge")
# Format y-axis scale and set y-axis label
cropDmgChart = cropDmgChart + ylab("Frequency Count") 
# Set x-axis label
cropDmgChart = cropDmgChart + xlab("Event Type") 
# Rotate x-axis tick labels 
cropDmgChart = cropDmgChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
cropDmgChart = cropDmgChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))
cropDmgChart
dev.off()

```

### chart for top 20 events causing the most total economic damages
You can also embed plots, for example:
```{r totalDmgChart}
# Create chart
png("totaldmg.png")
totalDmgChart <- ggplot(table_total, aes(x=EVTYPE, y=PROPERTY_DAMAGES))
# Plot data as bar chart
totalDmgChart = totalDmgChart + geom_bar(stat="identity", aes(fill=EVTYPE), position="dodge")
# Format y-axis scale and set y-axis label
totalDmgChart = totalDmgChart + ylab("Frequency Count") 
# Set x-axis label
totalDmgChart = totalDmgChart + xlab("Event Type") 
# Rotate x-axis tick labels 
totalDmgChart = totalDmgChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
totalDmgChart = totalDmgChart + ggtitle("Top 20 US events causing the economic damages") + theme(plot.title = element_text(hjust = 0.5))
totalDmgChart
dev.off()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
